---
title: "Class11: Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Section 1. PDB

Question 1. Determine the percentage of structures solved by X-Ray and Electron Microscopy. From the website what proportion of structures are protein? 
```{r}
stats<- read.csv("Data Export Summary.csv", row.names = 1)
stats
```
```{r}
percent.by.method<-stats$Total/sum(stats$Total) * 100
names(percent.by.method)<- rownames(stats)
percent.by.method
```
```{r}
sum(stats$Proteins)/sum(stats$Total)*100
```
## Section 3. Introduction to Bio3D in R
First we have to load the package we want to use.
```{r}
library(bio3d)
```
Now we can use the functions from this bio3d package
```{r}
pdb <- read.pdb("1hsg")
```
```{r}
pdb$atom$resid
```
We want to select out the protein and drug only parts of these molecular PDB files.
```{r}
prot.inds<-atom.select(pdb, "protein")
prot.inds
```
```{r}
pdb$atom[prot.inds$atom,]
prot.pdb<-trim.pdb(pdb,prot.inds)
write.pdb(prot.pdb, file="protein.pdb")
```
Lets do the same thing for our drug
```{r}
lig.inds<-atom.select(pdb, "ligand")
lig.pdb<-trim.pdb(pdb, lig.inds)
write.pdb(lig.pdb, file= "ligand.pdb")
```



