---
title: "Class8: Introduction to Machine Learning for Bioinformatics"
author: "Amber Fossier"
date: "2/5/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##K-means clustering

```{r}
#Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)

```

```{r}
km<-kmeans(x, center=2, nstart=20)
km$size
```
cluster membership assignment vector (i.e. which cluster group my data... )
```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
```

```{r}
km$totss
```

hierarchicall clustering
```{r}
#First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
#  clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc
```
```{r}
#Our input is a distance matrix from the dist()
# function. Lets make sure we understand it first
dist_matrix <- dist(x)
dim(dist_matrix)
NULL
View(as.matrix(dist_matrix))
dim(x)

dim( as.matrix(dist_matrix))
# Note. symmetrical pairwise distance matrix
```
```{r}
plot(hc)
abline(h=6, col="red")
cutree(hc, h=6)
```

we can also use k=number of cluster groups I want with **cutree()**
```{r}
cutree(hc, k=2)
```

##Differently linkage methods
```{r}
#Using different hierarchical clustering methods
d<-dist_matrix
hc.complete <- hclust(d, method="complete")
plot(hc.complete)
hc.average  <- hclust(d, method="average")
plot(hc.average)
hc.single   <- hclust(d, method="single")
plot(hc.single)
```

Run hclust() on this made up overlapping data
```{r}
#Step 1. Generate some example data for clustering
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
#         (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```

```{r}
dist_matrix <- dist(x)
hc <- hclust(d = dist_matrix)
plot(hc)
abline(h=2, col="red")
abline(h=2.5, col="blue")
```
```{r}
group2<- cutree(hc, k=2)
group3<- cutree(hc, k=2.5)
plot(x, col=group2)
```


```{r}
mydata<-read.csv("https://tinyurl.com/expression-CSV", row.names=1)
head(t(mydata))
pca<-prcomp(t(mydata), scale=TRUE)
attributes(pca)
plot(pca$x[,1],pca$x[,2])
pca.var<-pca$sdev^2
pca.var.per<-round(pca.var/sum(pca.var)*100,1)
barplot(pca.var.per, main="Scree Plot",
        xlab="Principal Component", ylab="Percent Variation")
```

```{r}
 ## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```
```{r}
food <- read.csv("UK_foods.csv", row.names=1)
head(food)
barplot(as.matrix(food), beside=T, col=rainbow(nrow(food)))
pairs(food, col=rainbow(10), pch=16)

```

```{r}
pca<-prcomp(t(food))
summary(pca)
```

```{r}
# Plot PC1 vs PC2

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col=c("orange", "red", "blue", "darkgreen"))
```

###Examine the "loadings"
This will help us determine how the original variables (dimensions) contribute to our new PCs
```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```
```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)

```

```{r}
rna.data <- read.csv("expression.csv", row.names=1)
head(rna.data)
```

